*flippant* -- An R package for the automated analysis of dithionite scramblase

### 2015-10-16 inst/manuscript/manuscript.rmd
More carefully omit taking ownership where it is Menn-lab only;

# Introduction
Scramblases are proteins mediating the bidirectional, mass-action driven 
equilibration of lipids between the leaflets of lipid bilayers that constitute 
biological membranes [@waechter_flippases_2013]. Independent of energy 
equivalents such as ATP they facilitate this ``flipping'' by providing a 
mechanism of transition through the hydrophobic environment of membranes to 
polar lipid head groups, thus reducing the massive energetic cost that process 
would carry otherwise [@kornberg_inside-outside_1971]. The underlying molecular 
mechanisms remain largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified [@menon_specific_2000]. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, we speculate that lipid scrambling may, in fact, 
be an intrinsic property of G-protein coupled receptors in general 
[@goren_constitutive_2014].

To interrogate scramblases as a functional class of proteins, we have developed 
a biochemical assay suited to characterize scramblase activity of purified 
proteins and protein extracts based on the reconstitution of scramblases or 
candidate proteins into preformed synthetic liposomes that contain trace amounts
of nitrobenzoxadiazole (NBD) labeled fluorescent lipids [@chang_chemical_2004].
Upon addition of membrane-impermeable reducing agents such as dithionite to 
these liposomes, fluorophores in the external leaflet are quenched and fluorescence in scramblase-free liposomes accordingly quickly drops to approximately 50% of the 
initial value. However, for proteoliposomes reconstituted with proteins 
conferring a scramblase activity, fluorescence disappears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the bilayer renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the liposome membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014], the 
assay has also been used to characterize the stoichiometry of scramblase 
homo-multimers reconstituting into the proteoliposomes 
[@ploier_perturbed_submitted] and probe the ATP-dependence of flippases, 
molecules that under energy-consumption flip lipids against a concentration 
gradient between bilayer leaflets [@soupene_atp8a1_2008].

@@

#Case Study
*Retinitis pigmentosa* is a degenerative disease of the retina, the majority of 
cases of which is linked to mutant forms of the G-protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by our recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved when tested using the dithionite scramblase assay
among other techniques. However, the same set of experiments pointed at a defect 
in rhodopsin dimerization prior to insertion into liposomes, generating the new 
hypothesis that the mutations may cause disease by interfering with the highly 
ordered quarternary structure of the molecule in the retina 
[@ploier_perturbed_submitted].

>>>>

# Introduction
Scramblases are proteins mediating the bidirectional, mass-action driven 
equilibration of lipids between the leaflets of lipid bilayers that constitute 
biological membranes [@waechter_flippases_2013]. Independent of energy 
equivalents such as ATP they facilitate this ``flipping'' by providing a 
mechanism of transition through the hydrophobic environment of membranes to 
polar lipid head groups, thus reducing the massive energetic cost that process 
would carry otherwise [@kornberg_inside-outside_1971]. The underlying molecular 
mechanisms remain largely enigmatic [for a discussion including exceptions see @ernst_phospholipid_2015].
Proteins supporting flipping processes as central as the equilibration of the 
ubiquitous phosphoglycerolipids across the membrane of the endoplasmic 
reticulum, where they are synthesized in the cytoplasmic monolayer, have yet to 
be identified [@menon_specific_2000]. 
None the less, there has been much recent progress in scramblase identification 
and characterization. Proteins shown to have scramblase activity include 
FtsW [@mohammadi_identification_2011], opsin/rhodopsin
[@menon_opsin_2011;@goren_constitutive_2014], TMEM16 
[@malvezzi_ca2-dependent_2013], $\beta$ 2-adrenergic and adenosine A2A 
receptors [both @goren_constitutive_2014]. Extrapolating from the latter findings 
and the opsin/rhodopsin case, one may speculate that lipid scrambling may, in
fact, be an intrinsic property of G-protein coupled receptors in general 
[@goren_constitutive_2014].

To interrogate scramblases as a functional class of proteins, a biochemical
assay was developed suited to characterize scramblase activity of purified 
proteins and protein extracts based on the reconstitution of scramblases or 
candidate proteins into preformed synthetic liposomes that contain trace amounts
of nitrobenzoxadiazole (NBD) labeled fluorescent lipids [@chang_chemical_2004].
Upon addition of membrane-impermeable reducing agents such as dithionite to 
these liposomes, fluorophores in the external leaflet are quenched and fluorescence in scramblase-free liposomes accordingly quickly drops to approximately 50% of the 
initial value. However, for proteoliposomes reconstituted with proteins 
conferring a scramblase activity, fluorescence disappears entirely, as
fast scramblase-mediated equilibration between outer (exposed) and inner
(protected) leaflet of the bilayer renders all fluorescent lipids
accessible to reduction. By titration of the protein amount reconstituted into
the liposome membranes it is thus possible to characterize scramblases and probe
their activity. Beyond the identification of scramblase activities in crude 
membrane preparations [@chang_chemical_2004;@rajasekharan_flip-flop_2011] and 
functional validation of candidate scramblases [@menon_opsin_2011;@malvezzi_ca2-dependent_2013;@goren_constitutive_2014], the 
assay has also been used to characterize the stoichiometry of scramblase 
homo-multimers reconstituting into the proteoliposomes 
[@ploier_perturbed_submitted] and probe the ATP-dependence of flippases, 
molecules that under energy-consumption flip lipids against a concentration 
gradient between bilayer leaflets [@soupene_atp8a1_2008].

@@

#Case Study
*Retinitis pigmentosa* is a degenerative disease of the retina, the majority of 
cases of which is linked to mutant forms of the G-protein coupled receptor 
rhodopsin [@mendes_mechanisms_2005]. Motivated by the recent discovery that 
opsin/rhodopsin has scramblase activity [@menon_opsin_2011;@goren_constitutive_2014],
we hypothesized that a class of enigmatic rhodopsin mutations, which are known
to cause *retinitis pigmentosa* yet do not display the commonly associated 
molecular phenotypes of impaired binding of retinal, folding and/or trafficking or 
transducin activation, may in fact be impacted in their scramblase activity.
This hypothesis was disproved when tested using the dithionite scramblase assay
among other techniques. However, the same set of experiments pointed at a defect 
in rhodopsin dimerization prior to insertion into liposomes, generating the new 
hypothesis that the mutations may cause disease by interfering with the highly 
ordered quarternary structure of the molecule in the retina 
[@ploier_perturbed_submitted].
